---
title: "Project Classification"
author: "Blaine Farrell"
date: "2022-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Some codes for Package
install.packages("rmarkdown");   
install.packages("tidyverse");
install.packages("ggplot2");
install.packages("RColorBrewer");
install.packages("kableExtra");
install.packages("papeR");
install.packages("epitab");

```{r, include=FALSE}
library(tidyverse);
library(ggplot2);
library(RColorBrewer);
library(MASS);
library(knitr);
library(kableExtra);
library(RColorBrewer);
library("papeR");
library(epitab);
library(caret);
myPalette <- brewer.pal(8, "Set2") 
```


##### Load in the Data and Preparation #####

```{r}
#nflreadr::.clear_cache()
library(gsisdecoder)
library(nflfastR)
library(dplyr, warn.conflicts = FALSE)
ids <- nflfastR::fast_scraper_schedules(2017:2019) %>%
  dplyr::filter(game_type == "SB") %>%
  dplyr::pull(game_id)
pbp <- nflfastR::build_nflfastR_pbp(ids)
```


```{r}
pbp <- nflfastR::load_pbp(2018:2020)
head(pbp)
```



```{r}
keep_col = c('home_team','home_coach','season','away_team','away_coach','posteam','defteam',
             'yards_gained','shotgun','no_huddle','qb_dropback','qb_scramble',
             'pass_location','run_location','run_gap','field_goal_result',
             'kick_distance','two_point_conv_result','timeout','timeout_team',
             'td_team','total_home_score','total_away_score','total_home_epa',
             'total_away_epa','total_home_rush_epa','total_away_rush_epa',
             'total_home_pass_epa','total_away_pass_epa','home_wp','away_wp',
             'first_down_rush','first_down_pass','first_down_penalty',
             'third_down_converted','third_down_failed','fourth_down_converted',
             'fourth_down_failed','penalty','rush_attempt','pass_attempt','touchdown',
             'pass_touchdown','rush_touchdown','penalty_team',
             'penalty_yards','replay_or_challenge_result','play_type_nfl',
             'drive_play_count','drive_time_of_possession')
pbp_epa = subset(pbp, select = keep_col)
```

```{r}
#CHANGE LOCATION VARIABLES TO NUMERIC

pbp_epa$pass_location[pbp_epa$pass_location == "left"] <- -5
pbp_epa$pass_location[pbp_epa$pass_location == "middle"] <- 0
pbp_epa$pass_location[pbp_epa$pass_location == "right"] <- 5

pbp_epa$run_location[pbp_epa$run_location == "left"] <- -5
pbp_epa$run_location[pbp_epa$run_location == "middle"] <- 0
pbp_epa$run_location[pbp_epa$run_location == "right"] <- 5

pbp_epa$run_gap[pbp_epa$run_gap == "guard"] <- 0
pbp_epa$run_gap[pbp_epa$run_gap == "tackle"] <- 1
pbp_epa$run_gap[pbp_epa$run_gap == "end"] <- 2

pbp_epa$replay_or_challenge_result[pbp_epa$replay_or_challenge_result == "reversed"] <- 1
pbp_epa$replay_or_challenge_result[pbp_epa$replay_or_challenge_result == "upheld"] <- 0
pbp_epa$replay_or_challenge_result[pbp_epa$replay_or_challenge_result == "denied"] <- 0

pbp_epa$two_point_conv_result[pbp_epa$two_point_conv_result == "failure"] <- 0
pbp_epa$two_point_conv_result[pbp_epa$two_point_conv_result == "success"] <- 1

pbp_epa$field_goal_result[pbp_epa$field_goal_result == "missed"] <- 0
pbp_epa$field_goal_result[pbp_epa$field_goal_result == "blocked"] <- 0
pbp_epa$field_goal_result[pbp_epa$field_goal_result == "made"] <- 1


```


```{r}
#CONVERT TIME OF POSSESSION INTO SECONDS

pbp_epa$drive_time_of_possession1<-with(as.POSIXlt(pbp_epa$drive_time_of_possession,format='%M:%S'),60*min +sec)

head(pbp_epa)
```


##### Coaches Dataset #####

```{r}
coach_sum <- aggregate(select_if(pbp_epa, is.numeric),  
                       list(pbp_epa$home_coach,pbp_epa$season), 
                       FUN=sum, na.rm=TRUE, na.omit=TRUE, na.action=NULL
)
coach_sum$Group.1 <- paste(coach_sum$Group.1,coach_sum$Group.2,sep="_")

sum_col = c('Group.1','timeout','penalty','touchdown')
coach_sum = subset(coach_sum, select = sum_col)

head(coach_sum)
```

```{r}
coach_mean <- aggregate(select_if(pbp_epa, is.numeric), 
                        list(pbp_epa$home_coach,pbp_epa$season), 
                        FUN=mean, na.rm=TRUE, na.omit=TRUE, na.action=NULL)

#coach_mean = subset(coach_mean, select = -c())
remove_cols = c('timeout','penalty')
coach_mean = subset(coach_mean, select = !(names(coach_mean) %in% remove_cols)) 

```

```{r}
# CLEAN COACH MEAN
#coach_mean$pass_touchdown1 <- coach_mean$pass_touchdown / coach_mean$touchdown
#coach_mean$rush_touchdown1 <- coach_mean$rush_touchdown / coach_mean$touchdown
coach_mean$third_conversion1 <- coach_mean$third_down_converted / (coach_mean$third_down_converted + coach_mean$third_down_failed)
coach_mean$fourth_conversion1 <- coach_mean$fourth_down_converted / (coach_mean$fourth_down_converted + coach_mean$fourth_down_failed)
#coach_mean$first_down_rush1 <- coach_mean$first_down_rush / (coach_mean$first_down_rush + coach_mean$first_down_pass + coach_mean$first_down_penalty)
#coach_mean$first_down_pass1 <- coach_mean$first_down_pass / (coach_mean$first_down_rush + coach_mean$first_down_pass + coach_mean$first_down_penalty)
#coach_mean$first_down_penalty1 <- coach_mean$first_down_penalty / (coach_mean$first_down_rush + coach_mean$first_down_pass + coach_mean$first_down_penalty)

```


```{r}
remove <- c('pass_touchdown','rush_touchdown','third_down_converted',
            'third_down_failed','fourth_down_converted','fourth_down_failed',
            'first_down_pass','first_down_rush','first_down_penalty',
            'kick_distance','field_goal_attempt','total_away_epa','total_away_rush_epa',
            'total_away_pass_epa','away_wp','touchdown','Group.2','season')

coach_mean$Group.1 <- paste(coach_mean$Group.1,coach_mean$Group.2,sep="_")
coach_mean1 <- subset(coach_mean, select = !(names(coach_mean) %in% remove)) 

head(coach_mean1)
```

##### Final Dataset #####

```{r}
coach <- merge(coach_mean1, coach_sum, by="Group.1")
keep_columns <- c('Group.1','home_wp','yards_gained','shotgun','no_huddle','qb_dropback',
               'qb_scramble','rush_attempt','pass_attempt','penalty_yards',
               'drive_play_count','drive_time_of_possession1','third_conversion1',
               'fourth_conversion1','timeout','penalty')
coach = subset(coach, select = keep_columns)

head(coach)


write.csv(coach,"C:\\Users\\Admin\\OneDrive\\Documents\\CLASSES\\FALL 2022\\STT450\\Project\\coach_dataset.csv", row.names = FALSE)

```


##### Classification #####

```{r}
coach <- read.csv("coach_dataset.csv")

dim(coach)
```



```{r}
cor(coach[,-1],coach$home_wp)
```



```{r}
med<-median(coach$home_wp)
coach$wp01 <- ifelse(coach$home_wp>med, 1, 0)
head(coach)
```



Train and Test data Split:
```{r}
set.seed(100)

#use 50% of the data for train and 50% for test
sample <- sample(c(TRUE, FALSE), nrow(coach), replace=TRUE, prob=c(0.5,0.5))
train <- coach[sample, ]
test.year <- coach[!sample, ]
dim(test.year)

test.coach <- test.year$wp01
```


```{r}
library(MASS)
```


##### LDA Analysis #####

  ### ALL

```{r}
start.time <- Sys.time()

lda.fit <- lda(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = coach)
lda.class <- predict(lda.fit, test.year)$class
table(lda.class, test.coach)
mean(lda.class == test.coach)
1 - mean(lda.class == test.coach)

end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


  ### 50 50 

```{r}
start.time <- Sys.time()

lda.fit50 <- lda(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = train)
lda.class50 <- predict(lda.fit50, test.year)$class
table(lda.class50, test.coach)
mean(lda.class50 == test.coach)
1 - mean(lda.class50 == test.coach)



end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


##### QDA Analysis #####

  ### ALL

```{r}
start.time <- Sys.time()




qda.fit <- qda(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = coach)
qda.class <- predict(qda.fit, test.year)$class
table(qda.class, test.coach)
mean(qda.class == test.coach)
1 - mean(qda.class == test.coach)


end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


  ### 50 50


```{r}
start.time <- Sys.time()

qda.fit50 <- qda(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = train)
qda.class50 <- predict(qda.fit50, test.year)$class
table(qda.class50, test.coach)
mean(qda.class50 == test.coach)
1 - mean(qda.class50 == test.coach)


end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


##### Logistic Regression Analysis #####


  ### ALL

```{r}
start.time <- Sys.time()

glm.fits <- glm(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = coach,
               family=binomial)
glm.probs <- predict(glm.fits, test.year, type = "response")
glm.pred <- rep(0, length(test.coach))
glm.pred[glm.probs > .5] <- 1
table(glm.pred, test.coach)
acc <- (16+21)/52
acc


end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


  ### 50 50

```{r}
start.time <- Sys.time()

glm.fits50 <- glm(wp01 ~  yards_gained+shotgun+no_huddle+qb_dropback+
               qb_scramble+rush_attempt+pass_attempt+penalty_yards+
               drive_play_count+drive_time_of_possession1+third_conversion1+
               fourth_conversion1+timeout+penalty, data = train,
               family=binomial)
glm.probs50 <- predict(glm.fits50, test.year, type = "response")
glm.pred50 <- rep(0, length(test.coach))
glm.pred50[glm.probs50 > .5] <- 1
table(glm.pred50, test.coach)
acc <- (16+12)/52
acc


end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


##### KNN Analysis #####

  ### ALL



```{r}
library(class)

for (i in 1: 10)
{
  start.time <- Sys.time()

      train.test <- cbind(coach$yards_gained,coach$shotgun,coach$no_huddle,
               coach$qb_dropback,coach$qb_scramble,coach$rush_attempt,
               coach$pass_attempt,coach$penalty_yards,coach$drive_play_count,
               coach$drive_time_of_possession1,coach$third_conversion1,
               coach$fourth_conversion1,coach$timeout,coach$penalty)
      
      set.seed(1)
      knn.pred <- knn(train.test, train.test, coach$wp01, k = i)
      
      conf.matrix<-table(knn.pred, coach$wp01)
      print(i)
      print(mean(knn.pred == coach$wp01))
      print(sensitivity(conf.matrix))
      print(specificity(conf.matrix))
      
      
      
  end.time <- Sys.time()
  time.taken <- end.time - start.time
  print(time.taken)
}

```



  ### 50 50


```{r}
for (i in 1: 10)
{
  start.time <- Sys.time()

    train.X <- cbind(train$yards_gained,train$shotgun,train$no_huddle,
               train$qb_dropback,train$qb_scramble,train$rush_attempt,
               train$pass_attempt,train$penalty_yards,train$drive_play_count,
               train$drive_time_of_possession1,train$third_conversion1,
               train$fourth_conversion1,train$timeout,train$penalty) #[train, ]
    test.X <- cbind(test.year$yards_gained,test.year$shotgun,test.year$no_huddle,
               test.year$qb_dropback,test.year$qb_scramble,test.year$rush_attempt,
               test.year$pass_attempt,test.year$penalty_yards,test.year$drive_play_count,
               test.year$drive_time_of_possession1,test.year$third_conversion1,
               test.year$fourth_conversion1,test.year$timeout,test.year$penalty)

    train.wp01 <- train$wp01
  
    set.seed(1)
    knn.pred <- knn(train.X, test.X, train.wp01, k = i)

    conf.matrix50<-table(knn.pred, test.year$wp01)
    print(i)
    print(mean(knn.pred == test.year$wp01))
    print(sensitivity(conf.matrix50))
    print(specificity(conf.matrix50))
    
    
  end.time <- Sys.time()
  time.taken <- end.time - start.time
  print(time.taken)
}


```



